---
title: "Preprocessing"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 1
    number_sections: true
editor: visual
---

## Initial Setup {.unnumbered}

The provided R code installs and loads the "WDI" package, a tool for accessing World Bank data indicators. Additionally, it installs and loads the "remotes" package, a dependency required for installation.

```{r}
#install.packages("WDI",repos = "http://cran.us.r-project.org",dependencies=TRUE,eval=FALSE)
library(WDI)

#install.packages("remotes",repos = "http://cran.us.r-project.org",dependencies=TRUE,eval=FALSE)
library(remotes)

```

## Data Preparation {.unnumbered}

The objective is to obtain specific World Development Indicators (WDI) data for the indicators listed in the "indicator_codes" vector, focusing on countries specified in the "countries" vector. The time range for data extraction is set between the years 2018 and 2023.

The WDI function is utilized with parameters such as indicator codes, countries, and the specified time range. The resulting dataset, denoted as "world_bank_data_2," is then refined by selecting relevant columns and renaming them for clarity. The summary dataset, named "world_bank_data_summary," includes columns such as country, year, population, GDP per capita, age dependency ratio, inflation, unemployment rate, female labor force participation rate, and male labor force participation rate.

Finally, the kable function is employed to render the summary dataset in HTML format, and the kable_styling function is applied to enhance the visual presentation of the table.

```{r}
library(gt)
library(knitr)
library(kableExtra)
```

```{r}
indicator_codes <- c("SP.POP.TOTL","NY.GDP.PCAP.CD","SP.POP.DPND","FP.CPI.TOTL.ZG","SL.UEM.TOTL.ZS","SL.TLF.CACT.FE.ZS","SL.TLF.CACT.MA.ZS")
countries <-  c("TUR","UKR","EUU")
start_date <- 2018
end_date <- 2023
world_bank_data_2 <- WDI( indicator = indicator_codes,
                        country = countries,
                        start = start_date, 
                        end = end_date, 
                        extra = FALSE)

world_bank_data_summary <- world_bank_data_2[,c("country","year","SP.POP.TOTL","NY.GDP.PCAP.CD","SP.POP.DPND","FP.CPI.TOTL.ZG","SL.UEM.TOTL.ZS","SL.TLF.CACT.FE.ZS","SL.TLF.CACT.MA.ZS")]
colnames(world_bank_data_summary) <- c("Country","Year","Population","GdpPerCapita","AgeDependancyRatio","Inflation","UnemploymentRate","FemaleLaborRate","MaleLaborRate")

options(max.print = 15)

kable(world_bank_data_summary, "html") %>% kable_styling()
```

## Save as .rds File {.unnumbered}

In this R code snippet, the saveRDS function is used to save the dataset named "world_bank_data_summary" as an RDS (R Data Serialization) file named "world_bank_data.rds." Subsequently, the readRDS function is employed to read the saved RDS file back into R and store it in a variable named "world_bank_data." This process enables the preservation and retrieval of the dataset in a serialized format.

```{r}
saveRDS(world_bank_data_summary, file = "world_bank_data.rds")
world_bank_data <- readRDS("world_bank_data.rds")
```

## Preprocessed Data Link {.unnumbered}

Please find preprocessed [Data Link](https://github.com/pjournal/mef07g-astralprojection/blob/main/world_bank_data.rds).
